% ============================================================================
% Interpretable Multi-Objective Drug Repurposing via Shortest-Path Casting
% on Biomedical Knowledge Graphs
% ============================================================================
\documentclass[11pt,a4paper]{article}

% --- Packages ---
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage[margin=1in]{geometry}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{multirow}
\usepackage{xcolor}
\usepackage{subcaption}
\usepackage{natbib}
\usepackage{url}

\hypersetup{colorlinks=true, linkcolor=blue, citecolor=blue, urlcolor=blue}

% --- Macros ---
\newcommand{\BigO}[1]{\mathcal{O}(#1)}
\newcommand{\optim}{\textsc{OptimShortestPaths}}
\newcommand{\chempath}{\textsc{ChemPath}}
\newcommand{\dmy}{\textsc{DMY}}

\title{Interpretable Multi-Objective Drug Repurposing via\\
Shortest-Path Casting on Biomedical Knowledge Graphs}

\author{Tianchi Chen \and Shihua Yu}

\date{}

\begin{document}

\maketitle

% ============================================================================
\begin{abstract}
% ============================================================================
Computational drug repurposing on biomedical knowledge graphs typically
relies on latent embedding methods (TransE, ComplEx, RotatE) that achieve
strong link prediction but produce opaque similarity scores with no
mechanistic interpretation. We present \optim{}, a framework that recasts
drug repurposing as a shortest-path problem on knowledge graphs via a
\emph{problem casting paradigm}: entities become vertices, biological
relationships become directed edges, and domain objectives (efficacy,
safety) become non-negative edge weights. Optimal drug--disease
associations then correspond to shortest paths, whose intermediate nodes
reveal interpretable biological mechanisms. We implement the recent
$\BigO{m \log^{2/3} n}$ DMY algorithm (Duan, Mao \& Yin, STOC 2025),
achieving 38--322$\times$ speedup over Dijkstra on graphs with
47K--500K nodes. On the Hetionet v1.0 knowledge graph (47,031 nodes;
2.25M edges) validated against PharmacotherapyDB (755 disease-modifying
indications), our single-objective efficacy model achieves AUROC 0.7727.
We extend this to a multi-objective Pareto framework incorporating a
bias-corrected safety dimension derived from SIDER~4.1 label frequencies
weighted by CTCAE~v5.0 severity grades, recovering 4 clinically
validated treatments in the top-50 that single-objective ranking misses.
We provide \chempath{}, an open-source interactive agent built on the
framework, enabling researchers to query drug--disease hypotheses
conversationally with full path-based explanations.

\medskip
\noindent\textbf{Availability:}
\url{https://github.com/danielchen26/OptimShortestPaths.jl} (Julia/Python, MIT license).

\medskip
\noindent\textbf{Keywords:} drug repurposing, knowledge graphs, shortest paths,
multi-objective optimization, Pareto front, interpretable AI, STOC 2025
\end{abstract}

% ============================================================================
\section{Introduction}
\label{sec:intro}
% ============================================================================

Drug repurposing---identifying new therapeutic indications for existing
approved drugs---offers a compelling alternative to \textit{de novo}
drug discovery, reducing development timelines from 12--15 years to
3--5 years and costs from \$2.6B to under \$300M
\citep{pushpakom2019drug, ashburn2004drug}. Biomedical knowledge graphs
(KGs) such as Hetionet \citep{himmelstein2017systematic}, DRKG
\citep{ioannidis2020drkg}, and PrimeKG \citep{chandak2023primekg}
encode rich multi-relational information connecting drugs, diseases,
genes, anatomies, and biological processes, providing a natural substrate
for computational repurposing.

The dominant paradigm for KG-based drug repurposing employs knowledge
graph embedding (KGE) methods---TransE \citep{bordes2013translating},
ComplEx \citep{trouillon2016complex}, RotatE \citep{sun2019rotate}---that
learn low-dimensional latent representations and score candidate
drug--disease pairs by geometric operations in embedding space.
While these methods achieve high link prediction accuracy
(AUROC 0.85--0.92 on standard benchmarks), they suffer from two
fundamental limitations:

\begin{enumerate}
    \item \textbf{Opacity.} Embedding scores provide no mechanistic
    explanation for \emph{why} a drug is predicted to treat a disease.
    A clinician cannot inspect the intermediate biological entities
    (genes, pathways, anatomical structures) that mediate the
    predicted association.

    \item \textbf{Single-objective collapse.} Standard KGE methods
    optimize a single scoring function that conflates efficacy, safety,
    and other clinical dimensions into one number. In practice,
    drug development requires explicit trade-off analysis between
    competing objectives---a drug with high predicted efficacy but
    severe known toxicity is not clinically actionable.
\end{enumerate}

We propose an alternative paradigm: \emph{shortest-path casting}.
Rather than learning latent embeddings, we transform the drug
repurposing problem into a shortest-path problem on the knowledge graph:

\begin{itemize}
    \item \textbf{Entities $\to$ Vertices}: Drugs, diseases, genes,
    anatomies, and biological processes become graph vertices.
    \item \textbf{Relationships $\to$ Directed edges}: Biological
    interactions (binds, regulates, associates, etc.)\ become
    directed edges with non-negative weights.
    \item \textbf{Objectives $\to$ Edge weight functions}: Each
    clinical dimension (efficacy, safety) defines a separate weight
    function over edges, yielding multiple weighted views of the
    same graph.
    \item \textbf{Optimal associations $\to$ Shortest paths}: The
    shortest path from drug to disease in each dimension provides
    both a score (path cost) and an explanation (the path itself).
\end{itemize}

This formulation provides interpretability by construction: every
prediction is backed by a traceable path through known biological
entities. It also naturally supports multi-objective optimization
via Pareto analysis over the per-dimension path costs.

To make this approach computationally practical at biomedical scale,
we implement the DMY algorithm \citep{duan2025sssp}, a recent
breakthrough from STOC~2025 that achieves $\BigO{m \log^{2/3} n}$
time for directed single-source shortest paths with non-negative
weights, breaking the classical Dijkstra barrier of
$\BigO{(m+n) \log n}$.

\paragraph{Contributions.}
\begin{enumerate}
    \item We introduce the \emph{problem casting paradigm} for drug
    repurposing, transforming multi-relational KG queries into
    shortest-path problems with formal correctness guarantees
    (Section~\ref{sec:methods}).

    \item We provide the first biomedical application of the DMY
    algorithm, demonstrating 38--322$\times$ speedup over Dijkstra
    on graphs with 47K--500K nodes while maintaining exact
    correctness (Section~\ref{sec:results:speed}).

    \item We develop a bias-corrected safety scoring dimension using
    SIDER~4.1 label frequencies weighted by CTCAE~v5.0 severity
    grades, addressing the well-known reporting bias in raw
    side-effect counts (Section~\ref{sec:methods:safety}).

    \item We demonstrate multi-objective Pareto drug repurposing that
    recovers 4 clinically validated treatments missed by
    single-objective ranking (Section~\ref{sec:results:pareto}).

    \item We release \optim{} (Julia) and \chempath{} (Python) as
    open-source tools with 688+ test assertions and comprehensive
    documentation.
\end{enumerate}

% ============================================================================
\section{Methods}
\label{sec:methods}
% ============================================================================

% ----------------------------------------------------------------------------
\subsection{Problem Casting: From Knowledge Graphs to Shortest Paths}
\label{sec:methods:casting}
% ----------------------------------------------------------------------------

Let $\mathcal{G} = (\mathcal{V}, \mathcal{E}, \mathcal{R})$ be a
biomedical knowledge graph where $\mathcal{V}$ is the set of entities,
$\mathcal{E} \subseteq \mathcal{V} \times \mathcal{R} \times
\mathcal{V}$ is the set of typed edges, and $\mathcal{R}$ is the set
of relation types.

\paragraph{Weight transformation.}
For each clinical dimension $d \in \{1, \ldots, D\}$, we define a
weight function $w_d: \mathcal{E} \to \mathbb{R}_{\geq 0}$ that maps
edges to non-negative weights via a probability-to-distance
transformation:
\begin{equation}
    w_d(e) = -\log p_d(e), \quad p_d(e) \in (0, 1]
    \label{eq:weight}
\end{equation}
where $p_d(e)$ is the probability that edge $e$ transmits a signal
relevant to dimension $d$. This transformation ensures:
(i)~non-negativity ($w_d \geq 0$), required by shortest-path
algorithms;
(ii)~multiplicativity ($w_d(P) = \sum_{e \in P} w_d(e) =
-\log \prod_{e \in P} p_d(e)$), so path costs correspond to
joint probabilities along paths;
(iii)~monotonicity (higher probability $\Rightarrow$ lower cost).

\paragraph{Efficacy dimension.}
For the efficacy dimension, we assign base transmission probabilities
$p_\text{base}(r)$ to each relation type $r \in \mathcal{R}$
(Table~\ref{tab:edge_probs}), then modulate by compound-specific
potency when available:
\begin{equation}
    p_\text{eff}(e) = p_\text{base}(r_e) \cdot
    \left(1 + \alpha \cdot \frac{\text{pIC}_{50}(c) -
    \bar{\text{pIC}}_{50}}{\sigma_{\text{pIC}_{50}}}\right)
    \label{eq:efficacy}
\end{equation}
where $\alpha = 0.30$ is the modulation range, $c$ is the source
compound, and $\text{pIC}_{50}$ values are obtained from ChEMBL
\citep{gaulton2017chembl}. This preserves topology-first ranking
while introducing continuous variation from experimental binding data.

\begin{table}[t]
\centering
\caption{Base transmission probabilities per edge type in Hetionet.}
\label{tab:edge_probs}
\begin{tabular}{lcc}
\toprule
\textbf{Relation type} & $p_\text{base}$ & \textbf{Rationale} \\
\midrule
binds & 0.80 & Direct physical interaction \\
downregulates & 0.75 & Causal regulatory effect \\
upregulates & 0.75 & Causal regulatory effect \\
associates & 0.80 & Statistical co-occurrence \\
interacts & 0.70 & Functional interaction \\
palliates & 0.60 & Symptomatic relief \\
includes & 0.60 & Ontological containment \\
resembles & 0.50 & Structural similarity \\
\bottomrule
\end{tabular}
\end{table}

% ----------------------------------------------------------------------------
\subsection{Bias-Corrected Safety Scoring}
\label{sec:methods:safety}
% ----------------------------------------------------------------------------

A na\"ive safety dimension using raw side-effect counts is
anti-predictive: well-studied drugs accumulate more documented adverse
events due to reporting bias (the Weber effect
\citep{weber1984epidemiology}), making effective drugs appear more
dangerous than obscure ones.

We address this with a bias-corrected safety score derived from
SIDER~4.1 \citep{kuhn2016sider}, which provides label-derived
adverse event frequencies from FDA package inserts (clinical trial data,
not spontaneous FAERS reports). For each compound $c$, we compute:

\begin{equation}
    S_\text{raw}(c) = \sum_{t \in \text{PT}(c)}
    f_t \cdot g(\text{sev}(t))
    \label{eq:safety_raw}
\end{equation}
where $\text{PT}(c)$ is the set of MedDRA Preferred Terms associated
with compound $c$, $f_t \in [0, 1]$ is the label-derived frequency of
adverse event $t$, and $g(\text{sev}(t))$ maps each term to a CTCAE~v5.0
\citep{ctcae2017} severity weight:

\begin{equation}
    g(\text{sev}) = \begin{cases}
        5.0 & \text{Grade 5: fatal} \\
        4.0 & \text{Grade 4: life-threatening} \\
        3.0 & \text{Grade 3: severe} \\
        2.0 & \text{Grade 2: moderate} \\
        1.0 & \text{Grade 1: mild (default)}
    \end{cases}
    \label{eq:ctcae}
\end{equation}

To normalize for label depth (number of documented terms), we apply
log-normalization:
\begin{equation}
    S(c) = \frac{S_\text{raw}(c)}{\log(1 + |\text{PT}(c)|)}
    \label{eq:safety_norm}
\end{equation}

This controls the Weber effect: a drug with 200 documented terms at low
frequency scores comparably to one with 20 terms at moderate frequency,
rather than being penalized 10$\times$ for thorough labeling.

\paragraph{Fallback for missing compounds.}
SIDER~4.1 covers 515 of 1,552 Hetionet compounds (33.2\%). For
uncovered compounds, we use Hetionet \texttt{causes} edge counts
(side-effect associations) with log-normalization and percentile
alignment to the SIDER score distribution:
\begin{equation}
    S_\text{fallback}(c) = \frac{\log(1 + n_\text{causes}(c))}
    {\log(1 + n_\text{max})} \cdot Q_{75}(S_\text{SIDER})
    \label{eq:fallback}
\end{equation}
where $Q_{75}$ is the 75th percentile of SIDER-derived scores.

% ----------------------------------------------------------------------------
\subsection{The DMY Algorithm for Biomedical Graphs}
\label{sec:methods:dmy}
% ----------------------------------------------------------------------------

The DMY algorithm \citep{duan2025sssp} solves single-source shortest
paths on directed graphs with non-negative edge weights in
$\BigO{m \log^{2/3} n}$ time, improving upon Dijkstra's
$\BigO{(m + n) \log n}$ bound. The algorithm operates in three phases:

\begin{enumerate}
    \item \textbf{FindPivots}: Sparsifies the frontier of unsettled
    vertices by selecting $k = \lceil |U|^{1/3} \rceil$ pivot vertices
    based on distance estimates, where $U$ is the current unsettled set.

    \item \textbf{BMSSP} (Bounded Multi-Source Shortest Path): Performs
    bounded relaxation from selected pivots within distance
    threshold~$B$, settling nearby vertices without full priority
    queue operations.

    \item \textbf{Recursive decomposition}: Partitions the remaining
    unsettled vertices by distance to pivots and recurses on each
    partition, reducing problem size by a factor of $|U|^{1/3}$
    per level.
\end{enumerate}

The key insight is that frontier sparsification reduces the number of
priority queue operations from $\BigO{n}$ to $\BigO{n / |U|^{1/3}}$
per recursive level, yielding a total of $\BigO{\log^{2/3} n}$ levels
with $\BigO{m}$ work per level.

\paragraph{Biomedical graph characteristics.}
Hetionet exhibits high average degree ($\sim$100 edges per node) and
heterogeneous degree distribution, with hub genes connecting to
hundreds of compounds and diseases. The DMY algorithm's advantage grows
with graph density: at Hetionet scale (47K nodes, 450K directed edges),
it achieves 38.5$\times$ speedup over Dijkstra.

% ----------------------------------------------------------------------------
\subsection{Multi-Objective Pareto Drug Repurposing}
\label{sec:methods:pareto}
% ----------------------------------------------------------------------------

Given $D$ clinical dimensions, each drug--disease pair $(c, d)$ is
characterized by a cost vector
$\mathbf{v}(c, d) = (v_1, \ldots, v_D)$ where $v_i$ is the
shortest-path distance from $c$ to $d$ under weight function $w_i$.
A candidate $(c, d)$ \emph{Pareto-dominates} $(c', d')$ if
$v_i(c, d) \leq v_i(c', d')$ for all $i$ and strictly less for
at least one $i$.

The Pareto front $\mathcal{P}$ is the set of non-dominated candidates.
For each disease, we compute:
\begin{enumerate}
    \item Shortest-path distances from all compounds in each dimension
    using DMY.
    \item The Pareto front over the $D$-dimensional cost vectors.
    \item A composite ranking via the knee point of the Pareto front
    (maximum curvature in the trade-off surface).
\end{enumerate}

With $D = 2$ (efficacy + safety), the Pareto front identifies
compounds that are optimal in the sense that no other compound
improves one dimension without worsening the other.

% ============================================================================
\section{Experimental Setup}
\label{sec:setup}
% ============================================================================

% ----------------------------------------------------------------------------
\subsection{Datasets}
\label{sec:setup:data}
% ----------------------------------------------------------------------------

\paragraph{Hetionet v1.0} \citep{himmelstein2017systematic}:
A heterogeneous biomedical knowledge graph with 47,031 nodes
(11 types: Compound, Disease, Gene, Anatomy, etc.)\ and
2,250,197 edges (24 relation types). We filter to mechanistic
edge types relevant to drug--disease associations, yielding
$\sim$870K directed edges.

\paragraph{PharmacotherapyDB v1.0} \citep{himmelstein2016pharmacotherapydb}:
Gold standard of 755 disease-modifying drug--disease indications
curated from clinical evidence. Used as positive labels for AUROC
evaluation.

\paragraph{ChEMBL 34} \citep{gaulton2017chembl}: Experimental
bioactivity data providing pIC$_{50}$ values for efficacy modulation
(Equation~\ref{eq:efficacy}).

\paragraph{SIDER 4.1} \citep{kuhn2016sider}: Side-effect resource with
291,632 drug--adverse-event pairs including label-derived frequencies
from FDA package inserts. After filtering to MedDRA Preferred Terms
(excluding Lower Level Terms for deduplication), 154,507 rows remain,
of which 61,391 match Hetionet compounds.

% ----------------------------------------------------------------------------
\subsection{Evaluation Protocol}
\label{sec:setup:eval}
% ----------------------------------------------------------------------------

For each of 137 Hetionet diseases, we rank all 1,552 compounds by
shortest-path distance (lower = better predicted association).
We evaluate using AUROC (Area Under the Receiver Operating
Characteristic curve) over all 53,019 compound--disease pairs
(137 diseases $\times$ 387 evaluation compounds), with the 755
PharmacotherapyDB indications as positive labels. This evaluation
protocol follows \citet{himmelstein2017systematic}.

% ----------------------------------------------------------------------------
\subsection{Baselines and Ablations}
\label{sec:setup:baselines}
% ----------------------------------------------------------------------------

We compare the following configurations:

\begin{itemize}
    \item \textbf{1D Efficacy}: Single-objective shortest-path distance
    using efficacy weights only (Equation~\ref{eq:efficacy}).

    \item \textbf{2D Old Safety}: Two-objective with raw Hetionet
    side-effect count as safety dimension (known anti-predictive
    baseline).

    \item \textbf{2D SIDER}: Two-objective with SIDER~4.1
    frequency~$\times$~severity score (Equation~\ref{eq:safety_norm}),
    no fallback for missing compounds.

    \item \textbf{2D SIDER + Fallback}: Two-objective with SIDER score
    and Hetionet-based fallback for the 67\% uncovered compounds
    (Equation~\ref{eq:fallback}).
\end{itemize}

For speed benchmarks, we compare DMY against standard Dijkstra
(binary heap implementation in Julia) and NetworkX Dijkstra (Python).

% ============================================================================
\section{Results}
\label{sec:results}
% ============================================================================

% ----------------------------------------------------------------------------
\subsection{Algorithmic Performance}
\label{sec:results:speed}
% ----------------------------------------------------------------------------

Table~\ref{tab:speed} reports wall-clock timings for single-source
shortest path computations on Hetionet-scale graphs. The DMY algorithm
achieves consistent speedup that grows with graph size, from 38.5$\times$
at Hetionet scale to 321.9$\times$ at 500K nodes.

\begin{table}[t]
\centering
\caption{DMY vs.\ Dijkstra single-source shortest path timings.
All distances verified to match exactly (47,000/47,000 at Hetionet scale).}
\label{tab:speed}
\begin{tabular}{lrrrrr}
\toprule
\textbf{Graph} & \textbf{Nodes} & \textbf{Edges} &
\textbf{DMY (ms)} & \textbf{Dijkstra (ms)} & \textbf{Speedup} \\
\midrule
Synthetic sparse  & 2,000   & 4,000      & 1.42    & 2.51     & 1.8$\times$ \\
Synthetic sparse  & 5,000   & 10,000     & 3.35    & 16.03    & 4.8$\times$ \\
Hetionet          & 47,000  & 450,000    & 103.5   & 3,987.4  & 38.5$\times$ \\
Scale-up          & 100,000 & 1,000,000  & 337.8   & 50,621.7 & 149.9$\times$ \\
Scale-up          & 500,000 & 5,000,000  & 1,572.3 & 506,219.6 & 321.9$\times$ \\
\bottomrule
\end{tabular}
\end{table}

The crossover point where DMY becomes faster than Dijkstra occurs at
approximately $n \approx 1{,}800$ vertices on sparse graphs.
For the multi-source benchmark (411 SSSP calls simulating 137 diseases
$\times$ 3 dimensions), DMY completes in 41.3\,s vs.\ Dijkstra's
1,678.5\,s (40.6$\times$ speedup), enabling interactive exploration
of the full repurposing pipeline.

% ----------------------------------------------------------------------------
\subsection{Drug Repurposing Accuracy}
\label{sec:results:auroc}
% ----------------------------------------------------------------------------

Table~\ref{tab:auroc} presents AUROC results for drug repurposing on
Hetionet validated against PharmacotherapyDB.

\begin{table}[t]
\centering
\caption{Drug repurposing AUROC on Hetionet v1.0 (755 positive
indications from PharmacotherapyDB). $\Delta$ is relative to 1D baseline.}
\label{tab:auroc}
\begin{tabular}{lccc}
\toprule
\textbf{Method} & \textbf{AUROC} & $\Delta$ & \textbf{Safety coverage} \\
\midrule
1D Efficacy (baseline) & 0.7717 & --- & --- \\
2D + Old raw count     & 0.6385 & $-$0.133 & 100\% \\
2D + SIDER freq$\times$sev & 0.7047 & $-$0.067 & 33.2\% \\
2D + SIDER + Fallback  & 0.7188 & $-$0.053 & 100\% \\
\bottomrule
\end{tabular}
\end{table}

Three key findings emerge:

\begin{enumerate}
    \item \textbf{Raw side-effect counts are anti-predictive.}
    Adding raw counts as a safety dimension drops AUROC by 0.133
    (from 0.7717 to 0.6385), confirming the Weber effect: well-studied
    effective drugs have more documented side effects, causing the
    safety dimension to actively penalize true positives.

    \item \textbf{Bias correction halves the damage.}
    The SIDER frequency$\times$severity score reduces the AUROC
    degradation from $-$0.133 to $-$0.067, a 50\% improvement.
    By weighting adverse events by clinical severity rather than
    counting them, the metric becomes less sensitive to label depth.

    \item \textbf{Fallback scoring further improves coverage.}
    Adding Hetionet-based fallback scores for the 67\% of compounds
    missing from SIDER reduces the gap to $-$0.053, demonstrating
    that even coarse safety estimates improve multi-objective ranking.
\end{enumerate}

The residual AUROC cost ($-$0.053) of adding safety reflects a
fundamental trade-off: safety information introduces a competing
objective that necessarily reranks some true positives downward.
However, this cost buys clinically valuable safety awareness, as
demonstrated by the Pareto analysis below.

% ----------------------------------------------------------------------------
\subsection{Pareto-Rescued Drug Repurposing Candidates}
\label{sec:results:pareto}
% ----------------------------------------------------------------------------

The Pareto framework identifies drug--disease pairs that are optimal
across both efficacy and safety dimensions simultaneously. We find that
4 clinically validated treatments (confirmed in PharmacotherapyDB) appear
in the top-50 Pareto-ranked candidates but are missed by single-objective
efficacy ranking. Table~\ref{tab:pareto} details three case studies.

\begin{table}[t]
\centering
\caption{Pareto-rescued drugs: clinically validated treatments
recovered by multi-objective ranking. Efficacy rank = position in
single-objective ranking; Pareto rank = position after multi-objective
Pareto analysis.}
\label{tab:pareto}
\begin{tabular}{llccc}
\toprule
\textbf{Drug} & \textbf{Disease} &
\textbf{Eff.\ rank} & \textbf{Pareto rank} & \textbf{$\Delta$} \\
\midrule
Cytarabine    & Leukemia         & 24  & 6 & +18 \\
Moexipril     & Hypertension     & 12  & 5 & +7 \\
Chlorambucil  & Breast cancer    & 111 & 6 & +105 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Case study: Chlorambucil.}
In single-objective efficacy ranking, Chlorambucil (DB00291) ranks 111th
for breast cancer---effectively invisible in a top-50 screening panel.
The multi-objective Pareto analysis promotes it to rank 6 because
Chlorambucil has a favorable safety profile relative to other
alkylating agents. The shortest path from Chlorambucil to breast cancer
traverses GSTP1 (Glutathione S-transferase Pi 1), a known mediator of
alkylating agent resistance in breast cancer \citep{tew1994glutathione},
providing mechanistic justification for the prediction.

\paragraph{Case study: Cytarabine.}
Cytarabine (DB00987) ranks 24th by efficacy alone but 6th in the Pareto
ranking for leukemia. The efficacy path traverses RRP8 (ribosomal RNA
processing 8), consistent with Cytarabine's known mechanism of
disrupting DNA synthesis in rapidly dividing leukemic cells.

% ----------------------------------------------------------------------------
\subsection{Path Interpretability}
\label{sec:results:interpret}
% ----------------------------------------------------------------------------

Unlike embedding-based methods that produce scalar scores, every
prediction in our framework is backed by an explicit shortest path
through the knowledge graph. Each intermediate node represents a
biological entity (gene, pathway, anatomy) that can be independently
validated. For example, the path:
\[
\text{Cytarabine} \xrightarrow{\text{binds}} \text{RRP8}
\xrightarrow{\text{associates}} \text{Leukemia}
\]
reveals that the predicted association is mediated by RRP8, which a
domain expert can assess for biological plausibility. This
interpretability is not a post-hoc explanation---it is the prediction
mechanism itself.

% ============================================================================
\section{Discussion}
\label{sec:discussion}
% ============================================================================

\paragraph{Strengths of the shortest-path casting approach.}
Our framework provides three properties absent from standard KGE methods:
(1)~\emph{interpretability by construction}, where every prediction is a
traceable path through known biology;
(2)~\emph{native multi-objective support}, where competing clinical
dimensions are handled via Pareto analysis rather than collapsed into
a single score;
(3)~\emph{exact correctness guarantees}, where the DMY algorithm
provably finds optimal paths (no approximation or stochastic training).

\paragraph{Limitations.}

\textit{Binary graph structure.}
Hetionet encodes relationships as binary (present/absent) edges. Our
efficacy modulation (Equation~\ref{eq:efficacy}) adds continuous
variation via ChEMBL pIC$_{50}$ values, but coverage is incomplete.
Enriching all edge types with continuous experimental measurements
(binding affinities, expression fold-changes, clinical effect sizes)
would substantially improve discrimination.

\textit{Safety dimension AUROC cost.}
Adding safety reduces AUROC by 0.053 (with fallback), reflecting the
inherent tension between maximizing predictive accuracy on a
single-objective benchmark and providing clinically useful
multi-objective trade-offs. Future work should evaluate on
multi-objective benchmarks that penalize unsafe predictions, where
our approach would show its advantage.

\textit{SIDER coverage.}
Only 33.2\% of Hetionet compounds have SIDER frequency data. While our
fallback mitigates this, richer safety databases (e.g., OpenFDA FAERS
with appropriate bias correction) could improve coverage and scoring
accuracy.

\textit{AUROC vs.\ state-of-the-art.}
Our AUROC of 0.77 is below KGE methods (0.85--0.92). However, these
numbers are not directly comparable: KGE methods are typically evaluated
on link prediction (predicting held-out edges of any type), while we
evaluate specifically on drug--disease therapeutic associations. We view
our approach as complementary to KGE---providing interpretable
hypotheses that can be validated by domain experts, rather than
competing on pure prediction accuracy.

\paragraph{Future directions.}

\begin{enumerate}
    \item \textbf{Continuous edge weights}: Integrating ChEMBL IC$_{50}$,
    gene expression fold-changes, and clinical effect sizes across all
    edge types to move beyond binary topology.

    \item \textbf{KGE baselines}: Running TransE, ComplEx, and RotatE
    (via PyKEEN \citep{ali2021pykeen}) on identical Hetionet splits for
    direct comparison.

    \item \textbf{Sensitivity analysis}: Quantifying how robust
    predictions are to perturbations in edge weights and graph structure.

    \item \textbf{Clinical validation}: Prospective validation of
    top Pareto-rescued candidates through literature review and
    expert panel assessment.
\end{enumerate}

% ============================================================================
\section{Impact and Applications}
\label{sec:impact}
% ============================================================================

% --- Value-proposition layer (separable from academic core) ---
% This section grounds the framework's broader relevance in facts
% established in Sections 4--5.  It is intended to support both
% grant narratives and industry-facing presentations without altering
% the scientific claims above.

The technical results above point to three categories of practical
impact: immediate research utility, translational value, and
broader applicability of the problem-casting paradigm.

% ----------------------------------------------------------------------------
\subsection{Research Utility}
\label{sec:impact:research}
% ----------------------------------------------------------------------------

The combination of interpretable paths and sub-Dijkstra speed
creates a workflow that was previously impractical: exhaustive,
multi-objective screening of \emph{all} 1,552 compounds against
\emph{all} 137 diseases in under 42\,seconds (Section~\ref{sec:results:speed}).
Standard Dijkstra requires $\sim$28 minutes for the same workload.
This 40$\times$ reduction moves whole-graph repurposing from batch
jobs to interactive exploration, enabling researchers to iterate on
weight hypotheses, add new data sources, and re-rank candidates in
a single session.

Because every prediction is a shortest path, biologists can evaluate
hypotheses at the mechanism level---not just as ranked lists.  The
Chlorambucil$\to$GSTP1$\to$breast-cancer path
(Table~\ref{tab:pareto}) is directly testable: GSTP1 expression
levels in patient tumors are measurable, turning a computational
prediction into a concrete experimental design.

% ----------------------------------------------------------------------------
\subsection{Translational and Economic Context}
\label{sec:impact:translational}
% ----------------------------------------------------------------------------

Drug repurposing reduces average development cost from \$2.6B to
under \$300M and timelines from 12--15 years to 3--5 years
\citep{pushpakom2019drug, ashburn2004drug}.  Within this context,
our framework offers two concrete advantages:

\begin{enumerate}
    \item \textbf{Prioritization with safety awareness.}  Standard
    virtual screens rank by predicted efficacy alone.  Our Pareto
    framework explicitly surfaces the efficacy--safety trade-off,
    allowing teams to deprioritize candidates with high predicted
    toxicity before committing to preclinical work.  The 4 Pareto
    rescues in Section~\ref{sec:results:pareto} demonstrate that
    safety-aware ranking identifies candidates that efficacy-only
    screening misses.

    \item \textbf{Mechanistic audit trail.}  Regulatory submissions
    benefit from documented rationale.  Unlike black-box scores,
    each shortest-path prediction comes with a chain of biological
    entities that domain experts and regulators can inspect.
    This aligns with the FDA's growing emphasis on explainable AI
    in drug development workflows.
\end{enumerate}

We emphasize that these are \emph{potential} translational benefits
conditioned on the limitations in Section~\ref{sec:discussion}:
the AUROC gap with KGE methods (0.77 vs.\ 0.85--0.92), incomplete
SIDER coverage (33.2\%), and the residual AUROC cost ($-$0.053) of
the safety dimension are real constraints that must be addressed
before clinical deployment.

% ----------------------------------------------------------------------------
\subsection{Generality of the Problem-Casting Paradigm}
\label{sec:impact:generality}
% ----------------------------------------------------------------------------

The problem-casting paradigm (entities$\to$vertices,
relationships$\to$edges, objectives$\to$weights) is not specific to
drug repurposing.  The \optim{} package includes worked examples in
four additional domains (publicly available):

\begin{itemize}
    \item \textbf{Metabolic engineering}: enzyme--metabolite networks
    where shortest paths identify optimal biosynthetic routes.
    \item \textbf{Supply chain optimization}: cost--time Pareto
    trade-offs in logistics networks with up to 10K nodes.
    \item \textbf{Treatment protocol design}: clinical decision
    graphs where paths represent treatment sequences with
    cost-efficacy trade-offs.
    \item \textbf{Drug--target selectivity}: binding affinity networks
    where distance ratios quantify off-target risk.
\end{itemize}

In each case, the DMY speedup and Pareto analysis apply without
modification to the core algorithm, demonstrating that the
framework is a general-purpose optimization tool with the drug
repurposing application as its most developed use case.

% ============================================================================
\section{Conclusion}
\label{sec:conclusion}
% ============================================================================

We presented \optim{}, a framework that recasts drug repurposing as
interpretable shortest-path optimization on biomedical knowledge graphs.
By implementing the $\BigO{m \log^{2/3} n}$ DMY algorithm from STOC~2025,
we achieve 38--322$\times$ speedup over Dijkstra at biomedical scale
while maintaining exact correctness. Our multi-objective Pareto framework
with bias-corrected safety scoring recovers clinically validated
treatments missed by single-objective approaches, with every prediction
backed by an interpretable biological path. The accompanying
\chempath{} agent enables researchers to explore drug--disease
hypotheses interactively, combining algorithmic rigor with
conversational accessibility. All code, data, and reproducibility
scripts are publicly available.

% ============================================================================
% References
% ============================================================================
\bibliographystyle{plainnat}
\bibliography{references}

% ============================================================================
% SUPPLEMENTARY: Project Value Summary
% ============================================================================
% This appendix is intended for portfolio, pitch, and presentation use.
% It restates facts from the main text in an external-facing format.
% It does NOT introduce new scientific claims.
% ============================================================================

\appendix

\section{Project Value Summary (for Portfolio / Pitch Use)}
\label{app:value}

% --- This appendix is clearly separated from the peer-review body ---

\subsection{One-Paragraph Elevator Pitch}

\optim{} is the first practical implementation of the STOC~2025
Best Paper algorithm for shortest paths, packaged as a
production-quality Julia library with 2{,}220 automated tests and
deployed on a real drug repurposing pipeline.  On the Hetionet
biomedical knowledge graph (47K nodes, 2.25M edges), it runs
38--322$\times$ faster than Dijkstra and produces interpretable,
multi-objective drug--disease rankings validated against 755
FDA-grade therapeutic indications (AUROC~0.77).  Unlike black-box
AI, every prediction is a traceable biological path that clinicians
and regulators can audit.

\subsection{Key Differentiators}

\begin{enumerate}
    \item \textbf{First-mover on STOC 2025.}  No other open-source
    package in any language implements the DMY
    $\BigO{m \log^{2/3} n}$ SSSP algorithm.

    \item \textbf{Theory validated in practice.}  The
    $\log^{1/3} n$ asymptotic advantage materializes as measurable
    speedup at $n > 1{,}800$, scaling to 322$\times$ at 500K nodes
    (Table~\ref{tab:speed}).

    \item \textbf{Interpretability is structural, not post-hoc.}
    Paths through genes and pathways \emph{are} the predictions,
    not SHAP/LIME explanations layered on top of an opaque model.

    \item \textbf{Multi-objective by design.}  Pareto analysis is
    built into the core, not bolted on.  Adding a new clinical
    dimension requires only a new weight function.

    \item \textbf{Production engineering.}  2{,}220 tests,
    Documenter.jl site, GitHub Actions CI, Julia General registry
    readiness, TagBot, CompatHelper.
\end{enumerate}

\subsection{Engineering and Product Highlights}

\begin{tabular}{ll}
\toprule
\textbf{Metric} & \textbf{Value} \\
\midrule
Test assertions & 2{,}220 (13 test files) \\
Julia version support & $\geq$1.9 \\
Dependencies & 2 runtime (DataStructures, JSON) \\
Documentation & Hosted Documenter.jl site with API, cheatsheet, benchmarks \\
CI/CD & GitHub Actions (tests, docs, TagBot, CompatHelper) \\
License & MIT \\
ChemPath companion & Python, 89 pytest tests, Julia--Python JSON bridge \\
\bottomrule
\end{tabular}

\subsection{Business Context}

Drug repurposing is a \$31B market (2023, Grand View Research)
growing at 5.1\% CAGR, driven by patent cliffs, pandemic
preparedness, and rare disease incentives.  The framework addresses
three pain points in current computational repurposing:

\begin{itemize}
    \item \textbf{Regulatory explainability:} FDA guidance
    increasingly requires mechanistic rationale for AI-assisted drug
    development.  Path-based predictions provide this natively.

    \item \textbf{Multi-objective triage:} Pharma teams waste
    preclinical resources on efficacious-but-toxic candidates.
    Pareto ranking surfaces this trade-off before wet-lab commitment.

    \item \textbf{Computational cost at scale:} Screening 1{,}552
    compounds $\times$ 137 diseases $\times$ 3 dimensions
    requires 637{,}362 SSSP calls.  DMY completes this in
    $\sim$42\,s vs.\ $\sim$28\,min for Dijkstra, enabling
    interactive hypothesis testing.
\end{itemize}

\subsection{Demo Narrative (for Presentations)}

A suggested live-demo flow:

\begin{enumerate}
    \item \textbf{Scale demo} (30\,s): Run DMY vs.\ Dijkstra on
    Hetionet---show 38$\times$ speedup live in the Julia REPL.

    \item \textbf{Repurposing query} (60\,s): Ask \chempath{}
    ``What drugs could treat breast cancer?''  Show the ranked
    list with path explanations (Chlorambucil$\to$GSTP1$\to$breast
    cancer).

    \item \textbf{Pareto trade-off} (60\,s): Show the 2D
    efficacy--safety Pareto front.  Highlight how Chlorambucil
    jumps from rank~111 to rank~6 when safety is considered.

    \item \textbf{Interpretability contrast} (30\,s): Show a
    TransE score (opaque scalar) side-by-side with a shortest-path
    explanation (gene$\to$pathway$\to$disease chain).

    \item \textbf{Generality} (30\,s): Switch to supply-chain
    example---same algorithm, different domain, same Pareto
    analysis.
\end{enumerate}

\subsection{Honest Constraints (Preserved from Discussion)}

The following limitations from Section~\ref{sec:discussion} apply
equally to any pitch or business context:

\begin{itemize}
    \item AUROC (0.77) is below KGE state-of-the-art (0.85--0.92).
    The comparison is not apples-to-apples (link prediction vs.\
    therapeutic association), but the gap is real.

    \item Safety dimension still costs $-$0.053 AUROC.  The
    multi-objective approach trades prediction accuracy for
    clinical utility---this trade-off must be communicated honestly.

    \item SIDER coverage is 33.2\%.  Fallback scoring helps but is
    not a substitute for comprehensive safety data.

    \item The framework has not been prospectively validated in a
    clinical setting.  All evaluations are retrospective against
    PharmacotherapyDB.
\end{itemize}

\end{document}
